---
title: "Community contribution"
author: "Kenza Slaoui"
date: "2024-11-07"
output: html_document
---

```{r, warning = FALSE, results = 'hide'}
data(iris)
library(tidyverse)
library(gridExtra)
```

# Using different shapes, patterns, and sizes 

## geom_density

### Basic Plot
```{r, fig.height =5, fig.width=10, warning = FALSE}

g_none = ggplot(iris, aes(x = Petal.Length, color = Species)) +
  geom_density() +  
  labs(title = "Petal Length Distributions by Species",
       x = "Petal Length",
       y = "Density") +
  theme_light() 

g_none
```

### Improvement: Changing the size of the line and introducing different linetypes

```{r, fig.height =5, fig.width=10, warning = FALSE}
g_imp1 = ggplot(iris, aes(x = Petal.Length, color = Species, linetype = Species)) + # Different linetype by group
  geom_density(size = 1.5) +   # Increase the size
  labs(title = "Petal Length Distributions by Species",
       x = "Petal Length",
       y = "Density") +
  theme_light() +
  scale_linetype_manual(values = c("solid", "dashed", "dotted")) # Manually choose the linetypes

g_imp1
```

### Improvement: Changing the size of the line and introducing annotations

```{r, fig.height =5, fig.width=10, warning = FALSE}
g_imp2 = ggplot(iris, aes(x = Petal.Length, color = Species)) +
          geom_density(size = 1.5) +   # Increase the size
          labs(title = "Petal Length Distributions by Species",
               x = "Petal Length",
               y = "Density") +
          theme_light() +
          theme(legend.position = "none") +
          annotate("segment", x = 1.6, xend = 1.9, y = 2.3, yend = 2.5, colour = "black") + #Annotating
          annotate("text", x=2.1, y=2.55, label="setosa",color="black", size=4) +
          annotate("segment", x = 4.7, xend = 5, y = 0.8, yend = 1, colour = "black") +
          annotate("text", x=5.25, y=1.05, label="versicolor",color="black", size=4) +
          annotate("segment", x = 6, xend = 6.35, y = 0.5, yend = 0.7, colour = "black") +
          annotate("text", x=6.55, y=0.75, label="virginica",color="black", size=4)

g_imp2
```

```{r, fig.height =10, warning = FALSE}
grid.arrange(g_none, g_imp1, g_imp2)
```


## geom_point

### Improvement: Using different shapes

```{r, fig.height =10, fig.width=10, warning = FALSE}
g1 = ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
      geom_point(size = 2, aes(alpha = 0.8)) +
      labs(title = "Sepal Length vs Sepal Width",
           x = "Sepal Length",
           y = "Sepal Width") +
      theme_minimal() +
      scale_alpha(guide = 'none')
  
g2 = ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, shape = Species, alpha = 0.9)) +
      geom_point(size = 3) +
      labs(title = "Sepal Length vs Sepal Width",
           x = "Sepal Length",
           y = "Sepal Width") +
      theme_minimal() +
      scale_shape_manual(values = c(16, 17, 8)) + # Different shapes (circle, triangle, square)
      scale_alpha(guide = 'none')

grid.arrange(g1, g2, nrow = 2)
```


## geom_bar

### Improvement: Using different patterns without color

```{r, warning = FALSE}
library(ggplot2)
library(ggpattern)

ggplot(mpg, aes(x = class, pattern = as.factor(year))) + 
  geom_bar_pattern(position = "dodge",                  # Group bars side by side
                   aes(pattern_fill = as.factor(year)), # Fill pattern based on the year
                   stat = "count",                      # Count cars in each group
                   width = 0.7,                         # Adjust bar width
                   pattern_density = 0.1,               # Density of the pattern
                   pattern_angle = 45,                  # Angle of the pattern
                   fill = "white",                      # Set the background color of the bars to white
                   colour = "black") +                  # Color and define the outline of the bars
  scale_pattern_manual(values = c("crosshatch", "circle", "stripe", "diamond")) +  # Different patterns for each year 
  labs(title = "Grouped Bar Plot with Different Pattern Types by Year",
       x = "Car Class",
       y = "Count of Cars") +
  theme_minimal() 
```

### Improvement: Using different patterns with color

```{r, warning = FALSE}
ggplot(mpg, aes(x = class, fill = as.factor(year), pattern = as.factor(year))) + 
  geom_bar_pattern(position = "dodge",                  # Group bars side by side
                   aes(pattern_fill = as.factor(year)), # Fill pattern based on the year
                   stat = "count",                      # Count cars in each group
                   width = 0.7,                         # Adjust bar width
                   pattern_density = 0.1,               # Density of the pattern
                   pattern_angle = 45,                  # Angle of the pattern                      
                   colour = "black") +                  # Color and define the outline of the bars
  scale_pattern_manual(values = c("crosshatch", "circle", "stripe", "diamond")) +  # Different patterns for each year 
  labs(title = "Grouped Bar Plot with Different Pattern Types by Year",
       x = "Car Class",
       y = "Count of Cars") +
  scale_fill_manual(values = c("pink", "lightblue", "grey", "white")) +  # Custom colors
  theme_minimal()  

```


# Using Color

## Colors to Avoid

* Red and Green
* Yellow and Violet
* Blue and Green

## Choosing colors and Checking how they would be perceived by CB individuals

```{r echo = T, results = 'hide', warning = FALSE}
g3 = ggplot(iris, aes(y = Species, x = Petal.Length, fill = Species)) +
  geom_boxplot() +
  labs(title = "Petal Length Distribution by Species",
       y = "Species",
       x = "Petal Length") +
  theme_minimal()
```

### RColorBrewer and colorblindcheck

```{r, fig.height = 10, warning = FALSE}
library(RColorBrewer) 
library(colorblindcheck)

## All the palettes in RColorBrewer
display.brewer.all()

## Palettes that are color friendly in RColorBrewer
display.brewer.all(colorblindFriendly = TRUE)

## Let's check what a color would look like to different color blind people using the colorblindcheck package
Spectral = brewer.pal(n = 11, name = 'Spectral')
#display.brewer.pal(n = 9, name = 'Spectral')
palette_check(Spectral, plot = TRUE) #vector of hexadecimal color descriptions
```

### Make your own colors (Hexadecimal RGB color specification) and colorblindcheck

You can make a color by defining the intensity of the red, green and blue channels in #rrggbb.
Each is a combination of two hexadecimals (0 1 2 3 4 5 6 7 8 9 A B C D E F). The lowest intensity is 00 and the highest FF.

```{r, warning = FALSE}
## You can make your own palette and check
palette = c("#FFBF00","#F46D43","#ABDDA4","#D53E4F", "#66C2A5","#FF0000","#80FF00","#00FFFF")
palette_1 = c("#FFBF00","#80FF00","#F46D43","#FF0000","#ABDDA4","#D53E4F","#66C2A5","#00FFFF")
palette_check(palette, plot = TRUE)
```

### Include the colors in your plot 

```{r, warning = FALSE}
## Palettes that are color friendly in RColorBrewer
#display.brewer.all(colorblindFriendly = TRUE)

# add a palette to a plot
g3 + scale_fill_manual(values = palette) # manual(values = ..) for palettes constructed manually
g3 + scale_fill_brewer(palette = "RdBu") # brewer(palette = "..") for palettes in RColorBrewer
g3 + scale_fill_brewer(palette = "Dark2")
```

### Check how the plots would be perceived by CB individuals

* Non-Friendly color

```{r, warning = FALSE}
#How can we visualize our plots from the perspective of a color blind person 

library(dichromat)

g3 + scale_fill_manual(values = palette_1)

g3 + scale_fill_manual(values = dichromat(palette_1, type = "protan"))

g3 + scale_fill_manual(values = dichromat(palette_1, type = "deutan"))

g3 + scale_fill_manual(values = dichromat(palette_1, type = "tritan"))

```

* Friendly Color

```{r, warning = FALSE}
g3 + scale_fill_brewer(palette = "Dark2" )

g3 + scale_fill_manual(values = dichromat(RColorBrewer::brewer.pal(n=11, name = "Dark2"), type = "protan"))

g3 + scale_fill_manual(values = dichromat(RColorBrewer::brewer.pal(n=11, name = "Dark2"), type = "deutan"))

g3 + scale_fill_manual(values = dichromat(RColorBrewer::brewer.pal(n=11, name = "Dark2"), type = "tritan"))
```

## Other Palettes: dichromat, colorBlindness, viridis (Color Safe)


```{r, warning = FALSE}
library(paletteer)
library(colorBlindness)

paletteer_d("dichromat::BrowntoBlue_10")
paletteer_d("colorBlindness::paletteMartin")
paletteer_c("viridis::viridis", 30)

plot(paletteer_d("dichromat::BrowntoBlue_10"))
plot(paletteer_d("colorBlindness::paletteMartin"))
plot(paletteer_c("viridis::viridis", 30))

#colorBlindness::availablePalette()
#dichromat::colorschemes
```

* dichromat package: https://cran.r-project.org/web/packages/dichromat/dichromat.pdf
* colorBlindness package: https://cran.r-project.org/web/packages/colorBlindness/colorBlindness.pdf
* all color palettes documentation: https://r-charts.com/color-palettes/#discrete






